---
title: "80's New Romantics vs 90's Trance: An Analytical Comparison"
author: "Joy Crosbie"
date: "Spring 2021"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
    theme: yeti
    orientation: rows
    vertical_layout: fill
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(tidymodels)
library(ggdendro)
library(heatmaply)
library(spotifyr)
library(plotly)
library(kableExtra)
library(compmus)
library(patchwork)
library(stringr)
library(scales)
library(shinyLP)

get_conf_mat <- function(fit) {
  outcome <- .get_tune_outcome_names(fit)
  fit %>% 
    collect_predictions() %>% 
    conf_mat(truth = outcome, estimate = .pred_class)
}  

get_pr <- function(fit) {
  fit %>% 
    conf_mat_resampled() %>% 
    group_by(Prediction) %>% mutate(precision = Freq / sum(Freq)) %>% 
    group_by(Truth) %>% mutate(recall = Freq / sum(Freq)) %>% 
    ungroup() %>% filter(Prediction == Truth) %>% 
    select(class = Prediction, precision, recall)
}  

eighties <- get_playlist_audio_features("", "7hIxxRmPxdpBa6zuDUC1FM")
trance <- get_playlist_audio_features("", "3XKJmcxhdP8NXIoamf5zwc")
```


```{r create playlists}
playlists <-
  bind_rows(
    eighties %>% mutate(playlist = "New Romantic"),
    trance %>% mutate(playlist = "Trance")
  )
```




### Introduction to the playlists: **New Romantics vs. Trance** {data-commentary-width=700}
My corpus consists of a playlist containing 80's New Romantic songs and a playlist containing trance classics from the early 90's to the mid 2010's. I was inspired by a statement made by BBC Radio 2, referring to Blue Monday (New Order's 1983 hit), as "a crucial link between Seventies disco and the dance/house boom that took off at the end of the New Romantic." The repetitive use of synths and heavy drum beats is a known characteristic of both New Romantic and trance music, and lead me to **compare the two and see what the main differences are between the two genres**. To do this, I will examine:

* Pitch features
* Timbre features
* Tempo
* Other variables from the Spotify API, such as danceability and valence.

Although the synthesizer is common in both playlists, there are some obvious differences when it comes to the tones. The 'punchy' synth sounds from the 80's have become more complex chord progressions of sweeping strings in trance music. I believe this will also be apparent in the valence and keys of the compared genres, because of the "poppy" influences of the new-romantic style compared to the euphoric feeling of trance music. I also believe that the tempo of the trance tracks will be spread across a smaller range of beats per minute (BPM). I am curious to see the difference in the 'energy' measured by Spotify, as I believe both genres contain a high amount of energy, but the energy they contain is different, as described above.

Typical examples of interesting tracks from the 80's playlist would be:

* Blue Monday - New Order
* Fade To Grey - Visage. 

Both tracks contain a basic chord progression in a minor key played on a synthesizer repeated throughout the song, accompanied by a similar drum and high hat beat. 

Typical examples of interesting tracks from the trance playlist would be:

* Adagio for Strings - Tiësto
* Carte Blanche - Veracocha. 

Both songs contain well known riffs and are comprised of a musical form that distinctly builds tension to either end in "peaks" or "drops". Both songs are also in minor keys, a common characteristic of trance music.


***

<iframe src="https://open.spotify.com/embed/playlist/3XKJmcxhdP8NXIoamf5zwc" width="300" height="380" frameborder="0" allowtransparency="true" allow="encrypted-media"></iframe>

<iframe src="https://open.spotify.com/embed/playlist/7hIxxRmPxdpBa6zuDUC1FM" width="300" height="380" frameborder="0" allowtransparency="true" allow="encrypted-media"></iframe>